---
title: Gettin’ Paid
author: Simon Naitram
date: '2018-07-16'
slug: gettin-paid
categories: []
tags: []
type: "post"
image:
description:
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(reshape2)
library(ggtech)
```

Salaries in Barbados have reached an unsettling sour spot. Salaries are too low compared to the prices we have to pay; we feel very poor when we go to the supermarket. And yet, salaries in Barbados are too high for our exports to be competitive. We produce goods at too high prices for there to be great demand for them internationally. **Too low for comfort, too high for success. No bueno.**

At the height of the early '90s crisis, the public sector agreed to a 9% salary cut to save the economy.^1^ Despite the furore it created, this plan worked. Today, we acknowledge the genius of the strategy, but we subsequently implemented rules that means it could never be used again. In the aftermath of the crisis, we agreed that salary increases should be tied directly to increases in productivity. 

Of course that didn't happen. Andrew Downes reported that between 1995 and 2003 *real salaries* increased by 13.9%.^4^ The IMF reported a 12.1% increase in *real salaries* of public sector employees from 1992 to 2003. A *real salary increase* is defined as when your salary increased more than the prices in the store increased. 

```{r changes, echo=FALSE, message=FALSE,  warning=FALSE, fig.width=5,fig.height=3, fig.align='center'}

data <- data.frame(matrix(NA, ncol=0,nrow=26))
data$years <- seq(1991,2016,1)
data$Downes <- ifelse(data$years==1995,0,ifelse(data$years==2003, 13.9, NA))
data$IMF <- ifelse(data$years==1992,0,ifelse(data$years==2003, 12.1, NA))
data$prod <- ifelse(data$years==1991,0,ifelse(data$years==2016, 12.7, NA))

data <- melt(data, id.vars = "years")
data <- na.omit(data)

ggplot(data, aes(x=years, y=value, color=variable)) + 
  geom_point(size=1.5) + geom_line(size=1.2, linetype="dashed") +
  theme_tech(theme="google") +
  scale_color_tech(theme="google")+
  scale_fill_tech(theme="google")+
  theme(legend.position="none") +
    labs(x="Year", y="% Growth", title="Wages and Productivity", 
         subtitle="The Tortoise and the Hare")

```


Downes (2004) reported a 13.9% increase in the real wage between 1995 and 2003 alone. The IMF reported a 12.1% increase from 1992 to 2003 in the public sector. It is unlikely wages have fallen in the 14 years after. US productivity grew 47% from 1991 to 2016 while real wages increased 30%, roughly a 64% ratio. The least conservative estimate of Barbados productivity growth is 12.7% from 1991 to 2016, while real wage growth had already exceeded that in the period to 2003.


In the height of the early 1990's crisis, wages were cut by about 9% (Henry and Miller, "Institutions versus Policies: A Tale of Two Islands" 2009). This entailed much negotiation, public demonstrations, legal challenges, cost much political capital, led to the ousting of a government, but most of all, **it worked**. Barbadians widely acknowledge the genius of the plan today, but in the moment it was a great sacrifice that had to be made. In the aftermath of the crisis, it was agreed that wage increases would be tied directly to productivity increases.

The nominal wage we receive is the dollar figure on the cheque we get every week or month. Sometimes this has to increase just because prices increase. If your salary increases at exactly the rate that prices are incresing, then we say there's no *real wage* growth. If your salary increases faster than prices have increased, then we say that your *real wage* has grown. Your real wage is your nominal wage adjusted for inflation. Your real wage measures your purchasing power: how many goods/services you can actually buy with your salary.

Clever though it sounded, there were two flaws in this plan. The first is the obvious: do you really think we were going to follow it??? Nope. The second is less obvious. Even if we followed it, our productivity growth does not stand in isolation. Because we have pegged our dollar to the US dollar, we have to keep pace with them. 

There was a flaw in this plan, even if this plan were followed to the letter---the peg to the USD. To maintain international competitiveness, the ratio of real wage growth to productivity growth in Barbados would have to equal the US ratio. The complete opposite has happened. Downes (2004) reported a 13.9% increase in the real wage between 1995 and 2003 alone.
The IMF reported a 12.1% increase from 1992 to 2003 in the public sector. It is unlikely wages have fallen in the 14 years after. US productivity grew 47% from 1991 to 2016 while real wages increased 30%, roughly a 64% ratio. The least conservative estimate of Barbados productivity growth is 12.7% from 1991 to 2016, while real wage growth had already exceeded that in the period to 2003.

Why are wages so important? All inputs into the local production process are either imported or priced internationally---except labour. If Barbadian goods and services are not internationally competitive, then either the production processes are inefficient or technologically deficient, or the real wage is onerously high. Without exchange rate flexibility, we have to closely monitor the real wage.

Guess what, apart from all the taxes, one huge reason prices in the country are so high is because we get paid too much. So lowering wages would also lower prices. 

So how do we make Barbadians richer without just arbitrarily increasing people's wages? Sorry to sound like a broken record, but yes, the answer is productivity. Let me tie the concepts of a few previous posts together.

Increased public investment has the effect of raising the returns to private investment, and automatically leads to higher private investment. Higher private investment increases the marginal returns to labour, meaning that wages go up and more people get hired. It's a straightforward economic relationship---increase public investment and workers will become more productive so that they deserve higher wages.

> Let me give you a counter-example of how this works, since that's what Barbados has a lot of now. The government refused to invest in upgrading, maintaining or fixing the south coast sewerage problem. That meant there was a decline in public investment. The sewerage stank. It meant that no one wanted to do business on the south coast, workers got sick, and businesses were forced to close. The returns to the investments these businesses had made were shrinking rapidly, and some even made huge losses. When businesses closed, reduced their hours, or just didn't get any business, it meant that workers had to be sent home or their salaries cut. 

Amazingly, we can reverse-engineer this process. Instead, if we massively *increase* public investment, it will increase the returns to private investment. Businesses will install new technology, start new companies, open new branches. This will lead to higher wages as we get fancier jobs, and more employment as they will need more workers. Importantly, this also increases the returns to our own education! We won't feel like we went to school for nothing!

In order for our lives to keep getting better, our purchasing power needs to keep increasing. That is, we need to be able to buy more with our salaries for us to say, in an economic sense, that we're better off.

Asking for a raise sounds like an easy solution. If businesses all over Barbados simply walked into work one day and raised everyone's wages, what would happen? Let's go through the thought experiment. 


---

1. Henry and Miller, "Institutions versus Policies: A Tale of Two Islands" 2009
2. The Barbados Labour Market: A Diagnostic Study and Reform Proposals, prepared for the Inter-American Development Bank.
3. I was still a toddler/child when all this happened, so if I've got any of the details incorrect from my reading, then feel free to let me know. 
4. Andrew Downes (find reference) Downes (2004)