---
title: Where Does Our Money Go?
author: Simon Naitram
date: '2018-09-01'
slug: where-does-our-money-go
categories: []
tags: []
type: "post"
image:
description: Whose Fault is it Anyway?
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(ggplot2)
library(reshape2)
library(directlabels)
library(ggtech)
library(dplyr)
library(waffle)
```

Barbados' average income per person has not grown since 2008. From 2008 to 2017, Barbados' average income---or national income--- shrank by 2.8%. Compare that to the period 1992 to 2007, during which Barbados' national income doubled. Had we kept up that growth rate, national income would be 43% higher than it was in 2017. This paints a very depressing picture.

```{r gni, echo=FALSE, message=FALSE,  warning=FALSE, fig.width=6,fig.height=4, fig.align='center'}

data <- read_excel("~/Box Sync/WSABE/gni.xlsx")

reg <- lm(log(gni)~poly(year,1), data=data[which(data$year<2008),])
#reg <- loess(gni~year, control=loess.control(surface="direct"), data=data[which(data$year<2008),])
new <- data.frame(year = data$year)
pred <- predict(reg, newdata=new)
data$pred <- exp(pred)
data <- melt(data, id.vars = "year")

ggplot() + 
  #geom_point(data=data[which(data$variable=="gni"),], aes(x=year, y=value), size=1.5, color="royalblue") +
  geom_line(data=data[which(data$variable=="gni"),], aes(x=year, y=value), size=1.2, color="royalblue") +
  geom_line(data=data[which(data$variable=="pred"),], aes(x=year, y=value), linetype="dashed", size=0.8, color="coral1") +
  #geom_hline(yintercept=15560, linetype="dashed") +
  theme_tech(theme="google") +
  scale_color_tech(theme="google")+
  scale_fill_tech(theme="google")+
  theme(legend.position="none") +
  annotate("text", x=2012, y=21000, label="Trend", color="coral1") + 
  annotate("text", x=2014, y=14000, label="Actual", color="royalblue") + 
    labs(x="Year", y="Dollars", title="A Lost Decade", 
         subtitle="National Income per Capita")

```

This makes me sad. If your income doesn't change but prices rise, you are poorer. Our country is poorer. More people are poor. We are poor. Poverty is depressing. 

But I am also angry. We have lost of decade of progress. We have been set back by 10 years. But who am I angry with? Who *should* I be angry with? 








The only way for us to dig ourselves out of this hole is to invest in new technology. This investment has to happen in every aspect of our lives. It is clear that the government failed us by refusing to invest. But is that it? Are we simply the victims of poor leadership? 

**We are at fault.** In 2007, Barbados' private capital stock---the sum of all our business investment over time---was estimated to be 62% of the public capital stock. That ratio is the 152nd lowest in the world out of 168 countries. It has nothing to do with size either---the chart below shows no systematic link. 

```{r stock, echo=FALSE, message=FALSE,  warning=FALSE, fig.width=6,fig.height=4, fig.align='center'}
stock <- read_excel("~/Box Sync/WSABE/priv_gov in 2007.xlsx")
stock$BB <- ifelse(stock$country=="Barbados",1,0)
stock$priv_gov[which(stock$priv_gov>10)] <- NA

ggplot(stock, aes(x=lngdp, y=priv_gov)) + 
    geom_point(aes(color=factor(BB), size=factor(BB), alpha=factor(BB))) + 
    theme_tech(theme="google") +
  scale_fill_tech(theme="google") +
  theme(legend.position="none") + 
  #scale_alpha(range = c(0.5, 1)) + 
  scale_colour_manual(values = c("grey", "blue")) +
  scale_alpha_manual(values=c(0.6,0.85)) +
  labs(subtitle="Sorted by Gross National Income", 
       x="Size (GNI)", 
       y="Private/Public Capital Ratio", 
       title="Private versus Public Capital Stock") +
  annotate("segment", y = 7, yend = 0.9, x = 8.14, xend = 8.14, colour = "palevioletred1", size=1, alpha=0.9, arrow=arrow()) + 
  annotate("text", y=7.5, x=8.14, label="Barbados", color="palevioletred1")


```




There is compelling evidence to suggest that Barbados' business sector has severely skimped on new investment. They too have failed to upgrade, to invest. They have refused to take us into the 21st century. We live in an archaic society, unfit for purpose. 

This is especially disastrous for tourism. For years we've heard the alarm sounded on Barbados' ageing hotels and the disappearing beaches. Even as Barbados' hotels and beaches deteriorate, there has been no widespread investment drive to improve and expand them. There has been no substantial investment in Barbados.

**But is it entirely their fault?**

From 1992 to 2007, Barbados saved 13.6% of it's national income each year. That's a bit below the Latin America & Caribbean average of 18.6%. It's well below the world average of 23.1% for the period. So to begin with, we've undersaved. Undersaving means that, as a country, we haven't put aside enough money for businesses to invest with. 

But 13.6% isn't bad. So where did that money go? Well most of our savings go to the banks. So the question is, what did the banks do with our money? The answer is found in the chart below---nothing.

```{r switch, echo=FALSE, message=FALSE,  warning=FALSE, fig.width=7,fig.height=5, fig.align='center'}
switch <- read_excel("~/Box Sync/WSABE/productive credit.xlsx")
switch <- melt(switch, id.vars=c("date","year"))
#money <- money %>% group_by(year) %>% summarise(ratio=mean(ratio, na.rm=T))

#plot barchart

ggplot(switch, aes(x=date, y=value, fill=variable)) + 
  geom_area() +
  scale_fill_manual(values=c("#FFBF03","#00B723")) +
  theme_tech(theme="google") +
  theme(legend.position="none") + 
  annotate("text", x=switch$date[160], y=0.75, label="Productive", size=6) +
  annotate("text", x=switch$date[160], y=0.25, label="Non-Productive", size=6, color="white") +
  annotate("text", x=c(switch$date[15],switch$date[327]), y=c(1.05-switch$value[1],1.05-switch$value[342]), label=c(round(switch$value[1],2),round(switch$value[342],2)), size=5) +
    annotate("text", x=c(switch$date[15],switch$date[327]), y=c(switch$value[343]-0.05,switch$value[684]-0.05), label=c(round(switch$value[343],2),round(switch$value[684],2)), size=5, color="white") +
    labs(x="Year", y="Share of Bank Credit", title="Bank Lending by Type", 
         subtitle="Where does our money go?")
```

Commercial banks in Barbados have shifted the sands of the country beneath us. In 1990, 68% of our savings went to productive loans---loans to businesses who could generate economic activity. By 2018, they had completely transformed the landscape of Barbados, lending a ridiculously meagre 28% of our savings to businesses.

We decades we entrusted commericial banks with the responsibility of investing our money. It is clear that trust was deeply misplaced. Their selfish choices have led us to the edge of despair. 

Without doubt they will argue that Barbadians are not creative and are not entrepreneurial. They will argue that Barbadians have not created the business opportunities they can invest in. That is palpably false. I have already shown that business ideas in Barbados boomed over the last two decades. 

There simply has not been any support for new businesses to grow. The World Bank's Enterprise Survey in 2010 showed that the biggest obstacle for Barbadian firms was access to finance. 

**Our ideas are viewed as being `too risky' for banks to invest our money in.** Let the irrationality of that sink in. Barbados has become a twisted financial dystopia. Banks have chosen to completely minimise their risk at the expense of our economic growth. And there are no returns without risk. Hence why commercial banks can only offer us a meagre 0.05% return on our savings.

No longer. It is time we take matters into our hands. We need to find ways of getting our savings directly into the hands of businesses so that they can invest in new ideas. We need to provide the business guidance for ideas to flourish. We need to provide the environment so that competition prevails. 

But most importantly we need to get active. We need to invest. Today.

I'm not even talking about wild ideas. I even mean traditional business investment. Decades after Barbados established itself as a world-renowned tourist destination, we still have a truly limited number of hotels. Barbados' most beautiful beach (in my biased opinion)---Carlisle Bay--- lies overrun with bush and broken houses. There are two hotels on that stretch by my count. TWO. If your concern is that the beach must be available to the public, then demand the hotels be built across the road from the beach. If your concern is that is harms the integrity of historic Bridgetown, the do as countries all over the world have done and demand that the hotels be built in a style that fits with old Bridgetown. But build hotels. Our hotels are limited, and old, and tired. 


